Denotational Semantics for SSA with Weak Memory Operations

### 2023-10-18 pldi24/paper.tex
Functional tap
Rewriting skeleton

% Syntax
\newcommand{\letexpr}[3]{\ensuremath{\ms{let}\;#1 = #2\;\ms{in}\;#3}}
\newcommand{\letstmt}[3]{\ensuremath{\ms{let}\;#1 = #2; #3}}
\newcommand{\brb}[2]{\ms{br}\;#1\;#2}
\newcommand{\lbrb}[2]{\brb{\lbl{#1}}{#2}}
\newcommand{\ite}[3]{\ms{if}\;#1\;\{#2\}\;\ms{else}\;\{#3\}}
\newcommand{\ewhere}[2]{\ms{then}\;#1\;\ms{where}\;#2}
\newcommand{\where}[2]{#1\;\ms{where}\;#2}
\newcommand{\wbranch}[3]{#1(#2) \lto #3}
\newcommand{\lwbranch}[3]{\wbranch{\lbl{#1}}{#2}{#3}}
\newcommand{\bsplice}[3]{#1(#2)\;\{#3\}}
\newcommand{\lbsplice}[3]{\bsplice{\lbl{#1}}{#2}{#3}}
\newcommand{\csplits}[3]{#1 \mapsto #2;#3}
\newcommand{\cwk}[2]{#1 \mapsto #2}
\newcommand{\lwk}[2]{#1 \rightsquigarrow #2}
\newcommand{\tlin}[2]{\ms{lin}_{#2}(#1)}
\newcommand{\ltlin}[3]{\ms{lin}_{#3}(#1^{#2})}
\newcommand{\thyp}[3]{#1: {#2}^{#3}}
\newcommand{\lhyp}[3]{#1[#2](#3)}
\newcommand{\llhyp}[3]{\lhyp{\lbl{#1}}{#2}{#3}}
\newcommand{\rle}[1]{{\scriptsize\textsf{#1}}}
\newcommand{\taff}{\ms{a}}
\newcommand{\trel}{\ms{r}}
\newcommand{\tint}{\infty}
\newcommand{\hasty}[5]{#1 \vdash_{#2} #3: {#4}^{#5}}
\newcommand{\haslb}[3]{#1 \vdash #2 \rhd #3}
\newcommand{\lhaslb}[3]{#1 \vdash #2 \rhd #3}
\newcommand{\issubst}[3]{#1: #2 \mapsto #3}
\newcommand{\substcons}[3]{[{#2 \mapsto #3}]{#1}}

@@

\begin{figure}
  \begin{gather*}    
    \prftree[r]{\rle{br}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
      %TODO: consider allowing binding here, or should it be purely for lets?
      {\hasty{\Xi}{0}{a}{A}{\tint}}
      {\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}} \qquad
    \prftree[r]{\rle{ite}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\hasty{\Delta}{0}{e}{\mb{2}}{\tint}}
      {\haslb{\Xi}{s}{\ms{L}}}
      {\haslb{\Xi}{t}{\ms{L}}}
      {\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}} \\
    \prftree[r]{\rle{let-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
      {\hasty{\Xi}{1}{a}{A}{q}}
      {\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}} \\
    \prftree[r]{\rle{let2-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
      {\hasty{\Xi}{1}{a}{A \otimes B}{q}}
      {\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}} \\
    \prftree[r]{\rle{where}}
      {\haslb{\Gamma}{t}{\ms{L}}}
      {\lhaslb{\ms{L}}{L}{\ms{K}}}
      {\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}} \qquad
    \prftree[r]{\rle{nil-br}}
      {\lhaslb{\ms{L}}{\cdot}{\ms{L}}} \qquad
    \prftree[r]{\rle{cons-br}}
      {\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
      {\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}
      {\lhaslb{\ms{L}}{L, \lwbranch{\ell}{x: A}{t}}{\ms{K}}}
  \end{gather*}
  \caption{Typing rules for \isotopessa blocks}
  \Description{Typing rules for isotope-SSA blocks}
  \label{fig:ssa-block-typing}
\end{figure}

\begin{theorem}[Minimal Context] \
  For all \(a, A\), there exists \(\Delta\) such that \(\forall \Gamma, \hasty{\Gamma}{p}{a}{A}{q} \implies \cwk{\Gamma}{\Delta} \land \hasty{\Delta}{p}{a}{A}{q}\). If such a \(\Gamma\) exists, then \(\Delta\) is unique; in this case we call \(\Delta\) \textbf{minimal}.
\end{theorem}

\begin{figure}
  \begin{gather*}
    \prftree[r]{\rle{subst-nil}}
      {\issubst{\cdot}{\cdot}{\cdot}}
      \qquad
    \prftree[r]{\rle{subst-cons}}
      {\issubst{\gamma}{\Theta_\Gamma}{\Gamma}}
      {\hasty{\Theta_x}{0}{a}{A}{q} \text{ minimal}}
      {\tlin{\Theta_x}{q}}
      {\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}
      {\issubst{\substcons{\gamma}{x}{a}}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
  \end{gather*}
  \caption{Typing rules for \isotopessa substitutions}
  \Description{Typing rules for isotope-SSA substitutions}
  \label{fig:ssa-subst-typing}
\end{figure}

\begin{figure}
  \begin{gather*}
    \prftree[r]{\rle{subset-nil}}
      {\cdot \subseteq \cdot}
      \qquad
    \prftree[r]{\rle{subset-cons}}
      {\Gamma \subseteq \Delta}
      {\Gamma, \thyp{x}{A}{q} \subseteq \Delta, \thyp{x}{A}{q}}
      \qquad
    \prftree[r]{\rle{subset-drop}}
      {\Gamma \subseteq \Delta}
      {\Gamma \subseteq \Delta, \thyp{x}{A}{q}} 
    \\
    \prftree[r]{\rle{in-head}}
      {\thyp{x}{A}{q} \in \Gamma, \thyp{x}{A}{q}}
      \qquad
    \prftree[r]{\rle{in-cons}}
      {\thyp{x}{A}{q} \in \Gamma}
      {\thyp{x}{A}{q} \in \Gamma, \thyp{y}{B}{r}}
    \\
    \prftree[r]{\rle{var-head}}
      {x \in \Gamma, \thyp{x}{A}{q}}
      \qquad
    \prftree[r]{\rle{var-cons}}
      {x \in \Gamma}
      {x \in \Gamma, \thyp{y}{B}{r}} 
    \\
    \Gamma \cup \cdot = \Gamma \qquad
    \Gamma \cup (\Delta, \thyp{x}{A}{q}) = (\Gamma \cup \Delta), \thyp{x}{A}{q} \text{ if } x \notin \Gamma \qquad
    \Gamma \cup (\Delta, \thyp{x}{A}{q}) = \Gamma \cup \Delta \text{ otherwise} 
    % \\
    %   \cdot^\Gamma_\Delta = \cdot 
    %   \qquad
    %   (\Theta, \thyp{x}{A}{q})^\Gamma_\Delta
    %     = \Theta^\Gamma_\Delta, 
    %       \thyp{x}{A}{q} \text{ if } x \in \Delta \lor x \notin \Gamma
    %   \qquad
    %   (\Theta, \thyp{x}{A}{q})^\Gamma_\Delta = \Theta^\Gamma_\Delta \text{ otherwise}
    % \\
    %   \cdot^\Gamma_\Delta = \cdot 
    %   \qquad
    %   (\ms{L}, \llhyp{\ell}{\Theta}{A})^\Gamma_\Delta
    %   = \ms{L}^\Gamma_\Delta, \llhyp{\ell}{\Theta^\Gamma_\Delta}{A}
  \end{gather*}
  \caption{Manipulation of \isotopessa contexts}
  \Description{Manipulation of isotope-SSA contexts}
  \label{fig:context-manipulation}
\end{figure}

\TODO{rewrite union in rule-style?}

@@

\TODO{restriction of substitutions to contexts, further context manipulation}

\TODO{substitution is a RELATION}

\TODO{label substitution}

\TODO{lifting of substitutions to label substitutions}

\begin{theorem}[Substitution Weakening] \
  \TODO{this}
\end{theorem}

\begin{theorem}[Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then \(\haslb{\Theta}{[\gamma]t}{[\gamma]\ms{L}}\)
  \end{itemize}
\end{theorem}

\begin{figure}
  \begin{center}
    \TODO{this}
  \end{center}
  \caption{Grammar for \isotopessa blocks with holes}
  \Description{Grammar for isotope-SSA blocks with holes}
  \label{fig:blocks-with-holes-grammar}
\end{figure}

\begin{figure}
  \begin{center}
    \TODO{this}
  \end{center}
  \caption{Typing rules for \isotopessa blocks with holes}
  \Description{Typing rules for isotope-SSA blocks with holes}
  \label{fig:blocks-with-holes-typing}
\end{figure}

\TODO{rewriting}

@@

\begin{figure}
  \begin{center}
    \TODO{this}
  \end{center}
  \caption{Semantics for \isotopessa blocks with holes}
  \Description{Semantics for isotope-SSA blocks with holes}
  \label{fig:blocks-with-holes-semantics}
\end{figure}

\begin{theorem}[Rewriting] \
  \TODO{this}
\end{theorem}

>>>>

% Syntax
\newcommand{\letexpr}[3]{\ensuremath{\ms{let}\;#1 = #2\;\ms{in}\;#3}}
\newcommand{\letstmt}[3]{\ensuremath{\ms{let}\;#1 = #2; #3}}
\newcommand{\brb}[2]{\ms{br}\;#1\;#2}
\newcommand{\lbrb}[2]{\brb{\lbl{#1}}{#2}}
\newcommand{\ite}[3]{\ms{if}\;#1\;\{#2\}\;\ms{else}\;\{#3\}}
\newcommand{\ewhere}[2]{\ms{then}\;#1\;\ms{where}\;#2}
\newcommand{\where}[2]{#1\;\ms{where}\;#2}
\newcommand{\wbranch}[3]{#1(#2) \lto #3}
\newcommand{\lwbranch}[3]{\wbranch{\lbl{#1}}{#2}{#3}}
\newcommand{\bsplice}[3]{#1(#2)\;\{#3\}}
\newcommand{\lbsplice}[3]{\bsplice{\lbl{#1}}{#2}{#3}}
\newcommand{\csplits}[3]{#1 \mapsto #2;#3}
\newcommand{\cwk}[2]{#1 \mapsto #2}
\newcommand{\lwk}[2]{#1 \rightsquigarrow #2}
\newcommand{\tlin}[2]{\ms{lin}_{#2}(#1)}
\newcommand{\ltlin}[3]{\ms{lin}_{#3}(#1^{#2})}
\newcommand{\thyp}[3]{#1: {#2}^{#3}}
\newcommand{\lhyp}[3]{#1[#2](#3)}
\newcommand{\llhyp}[3]{\lhyp{\lbl{#1}}{#2}{#3}}
\newcommand{\rle}[1]{{\scriptsize\textsf{#1}}}
\newcommand{\taff}{\ms{a}}
\newcommand{\trel}{\ms{r}}
\newcommand{\tint}{\infty}
\newcommand{\hasty}[5]{#1 \vdash_{#2} #3: {#4}^{#5}}
\newcommand{\haslb}[3]{#1 \vdash #2 \rhd #3}
\newcommand{\lhaslb}[3]{#1 \vdash #2 \rhd #3}
\newcommand{\issubst}[3]{#1: #2 \mapsto #3}
\newcommand{\lbsubst}[3]{#1: #2 \rightsquigarrow #3}
\newcommand{\exprletsubst}[2]{{#1};{#2}}
\newcommand{\stmtletsubst}[2]{{#1};{#2}}
\newcommand{\mhole}[1]{{#1}^?}
\newcommand{\lhole}[1]{?#1}
\newcommand{\mhasty}[6]{#1;#2 \vdash_{#3} #4: {#5}^{#6}}
\newcommand{\mhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
\newcommand{\mlhaslb}[4]{#1;#2 \vdash #3 \rhd #4}
\newcommand{\tyhole}[5]{#1: #2 \mapsto_{#3} {#4}^{#5}}
\newcommand{\blkhole}[3]{#1: #2 \mapsto #3}
\newcommand{\cfghole}[3]{#1: #2 \mapsto #3}
\newcommand{\substctx}[2]{{#1}^{#2}}
\newcommand{\substlbs}[2]{{#1}^{#2}}
\newcommand{\restrictsubst}[2]{{#1}_{#2}}
\newcommand{\subsubst}[2]{#1 \subseteq #2}
\newcommand{\isrw}[3]{#1: #2 \mapsto #3}
\newcommand{\strictlbsubst}[3]{#1: #2 \rightsquigarrow_= #3}

@@

\begin{figure}
  \begin{gather*}    
    \prftree[r]{\rle{br}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
      {\hasty{\Xi}{0}{a}{A}{\tint}}
      {\haslb{\Gamma}{\lbrb{\ell}{a}}{\ms{L}}} \qquad
    \prftree[r]{\rle{ite}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\hasty{\Delta}{0}{e}{\mb{2}}{\tint}}
      {\haslb{\Xi}{s}{\ms{L}}}
      {\haslb{\Xi}{t}{\ms{L}}}
      {\haslb{\Gamma}{\ite{e}{s}{t}}{\ms{L}}} \\
    \prftree[r]{\rle{let-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\haslb{\Delta, \thyp{x}{A}{}}{t}{\ms{L}}}
      {\hasty{\Xi}{1}{a}{A}{q}}
      {\haslb{\Gamma}{\letstmt{x}{a}{t}}{\ms{L}}} \\
    \prftree[r]{\rle{let2-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\haslb{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{t}{\ms{L}}}
      {\hasty{\Xi}{1}{a}{A \otimes B}{q}}
      {\haslb{\Gamma}{\letstmt{(x, y)}{a}{t}}{\ms{L}}} \\
    \prftree[r]{\rle{where}}
      {\haslb{\Gamma}{t}{\ms{L}}}
      {\lhaslb{\ms{L}}{L}{\ms{K}}}
      {\haslb{\Gamma}{\ewhere{t}{L}}{\ms{K}}} \qquad
    \prftree[r]{\rle{nil-br}}
      {\lhaslb{\ms{L}}{\cdot}{\ms{L}}} \qquad
    \prftree[r]{\rle{cons-br}}
      {\lhaslb{\ms{L}}{L}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
      {\haslb{\Gamma, \thyp{x}{A}{}}{t}{\ms{L}}}
      {\lhaslb{\ms{L}}{L, \lwbranch{\ell}{x: A}{t}}{\ms{K}}}
  \end{gather*}
  \caption{Typing rules for \isotopessa blocks}
  \Description{Typing rules for isotope-SSA blocks}
  \label{fig:ssa-block-typing}
\end{figure}

\begin{figure}
  \begin{center}        
    \begingroup
    \renewcommand{\arraystretch}{1.5}
    \setlength{\tabcolsep}{2em}
    \begin{tabular}{rl}
        \multicolumn{1}{c}{Judgment} & \multicolumn{1}{c}{Meaning} \\ \hline
        \(\issubst{\gamma}{\Theta}{\Gamma}\) &
        ... \\
        \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) &
        ... \\
        \(\mhasty{H}{\Gamma}{p}{\mhole{a}}{A}{q}\) &
        ... \\
        \(\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}\) &
        ... \\
        \(\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}\) &
        ...
    \end{tabular}
    \endgroup
  \end{center}
  \caption{Metatheoretic typing judgements for \isotopessa}
  \Description{Metatheoretic typing judgements for isotope-SSA}
  \label{fig:ssa-meta-judgements}
\end{figure}

@@

\begin{figure}
  \begin{gather*}
    \boxed{\gamma: \ms{Var} \to \ms{Expr}}
    \\
    \prftree[r]{\rle{subst-nil}}
      {\issubst{\gamma}{\cdot}{\cdot}}
      \qquad
    \prftree[r]{\rle{subst-cons}}
      {\issubst{\gamma}{\Theta_\Gamma}{\Gamma}}
      {\hasty{\Theta_x}{0}{\gamma(x)}{A}{q}}
      {\tlin{\Theta_x}{q}}
      {\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}
      {\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
    \\
    \boxed{\mc{L}: \ms{Label} \to \ms{Expr} \to \ms{Block} \text{ s.t. for fresh } x, [a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)}
    \\
    \prftree[r]{\rle{lb-subst-nil}}
    {\lbsubst{\mc{L}}{\cdot}{\ms{K}}}
      \qquad
    \prftree[r]{\rle{lb-subst-cons}}
      {\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      {\haslb{\Gamma, \thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}}
      {\lbsubst{\mc{L}}{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}}}
    \\
    \prftree[r]{\rle{lb-strict}}
      {\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      {\ms{labels}(\ms{L}) = \ms{labels}(\ms{K})}
      {\strictlbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
  \end{gather*}
  \caption{Typing rules for \isotopessa substitutions. We assume that for all but finitely many \(x\), \(\lbl{\ell}\), \(\gamma(x) = x\) and \(\mc{L}(\lbl{\ell}, x) = \lbrb{\ell}{x}\).}
  \Description{Typing rules for isotope-SSA substitutions}
  \label{fig:ssa-subst-typing}
\end{figure}

\begin{figure}
  \begin{gather*}
    \subsubst{\gamma}{\gamma'} \iff 
      \forall x, \gamma(x) \in \{x, \gamma'(x)\}
    \\
    \restrictsubst{\gamma}{\Gamma}(x) 
      =
      (\ms{if}\; x \in \Gamma
      \;\ms{then}\;\gamma(x)
      \;\ms{else}\;x)
    \\
    \stmtletsubst{\substctx{\gamma}{\cdot}}{t} = t
    \qquad
    \stmtletsubst{\substctx{\gamma}{\Gamma, x: A^q}}{t} 
      = \letstmt{x}{\gamma(x)}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}
    \qquad
    \exprletsubst{\cdot}{e} = e
    \qquad
    \exprletsubst{\substctx{\gamma}{\Gamma, x: A^q}}{e} 
      = \letstmt{x}{\gamma(x)}{\substctx{\gamma}{\Gamma}}
    \\
    \substlbs{\gamma}{\cdot}(\lbl{\ell}, x) = \lbrb{\ell}{x}
    \qquad
    \substlbs{\gamma}{\ms{L}, \llhyp{\ell}{\Gamma}{A}}(\lbl{k}, x)
      = 
      (\ms{if}\; \lbl{k} = \lbl{\ell} 
      \;\ms{then}\;\stmtletsubst{\substctx{\gamma}{\Gamma}}{\lbrb{\ell}{x}}
      \;\ms{else}\;\substlbs{\gamma}{\ms{L}}(\lbl{\ell}, x))
    \\
    [\mc{L}](\lbrb{\ell}{a}) = \mc{L}(\lbl{\ell}, a)
    \qquad
    [\mc{L}]\ite{e}{s}{t} = \ite{e}{[\mc{L}]s}{[\mc{L}]t}
    \\
    [\mc{L}](\letstmt{x}{a}{t}) = (\letstmt{x}{a}{[\mc{L}]t})
    \qquad
    [\mc{L}](\letstmt{(x, y)}{a}{t}) = (\letstmt{(x, y)}{a}{[\mc{L}]t})
    \\
    [\mc{L}]\ewhere{t}{L} = \ewhere{[\mc{L}]t}{[\mc{L}]L}
    \\
    [\mc{L}]\cdot = \cdot
    \qquad
    [\mc{L}](\lwbranch{\ell}{x: A}{t})
    = (\lwbranch{\ell}{x: A}{[\mc{L}]t})
  \end{gather*}
  \caption{Operations on \isotopessa substitutions. In particular, note that for all \(\Gamma\), \(\subsubst{\restrictsubst{\gamma}{\Gamma}}{\gamma}\)}
  \Description{Operations on isotope-SSA substitutions}
  \label{fig:ssa-subst-ops}
\end{figure}

\begin{lemma}[Substitution Weakening] 
  If \(\csplits{\Gamma}{\Delta}{\Xi}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then there exist \(\Theta_\Delta\), \(\Theta_\Xi\) such that:
  \begin{itemize}
    \item \(\csplits{\Theta}{\Theta_\Delta}{\Theta_\Xi}\)
    \item \(\issubst{\gamma_\Delta}{\Theta_\Delta}{\Delta}\)
    \item \(\issubst{\gamma_\Xi}{\Theta_\Xi}{\Xi}\)
  \end{itemize}
  Furthermore, if \(\subsubst{\gamma}{\gamma'}\) and \(\issubst{\gamma}{\Theta}{\Gamma}\), then \(\issubst{\gamma'}{\Theta}{\Gamma}\).
\end{lemma}

\begin{theorem}[Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\)
    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), \(\strictlbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}\), and then \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\)
  \end{itemize}
\end{theorem}

\begin{theorem}[Label Substitution] 
  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), if \(\haslb{\Gamma}{t}{\ms{L}}\), then \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
\end{theorem}

\begin{figure}
  \begin{grammar}
    <\(\mhole{a}, \mhole{b}, \mhole{c}, \mhole{e}\)> ::= \(\lhole{X}\)
    \;|\; \(x\) 
    \;|\; \(f\;\mhole{a}\)
    \;|\; \((\mhole{a}, \mhole{b})\) 
    \;|\; \(()\) 
    \;|\; \(\ctt\) 
    \;|\; \(\cff\)
    \;|\; \(\letexpr{x}{\mhole{a}}{\mhole{e}}\)
    \;|\; \(\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}\)
    \alt \(\lbsplice{\ell}{x: A}{\mhole{t}}\)

    <\(\mhole{s}, \mhole{t}\)> ::= \(\lhole{T}\) 
    \;|\; \(\lbrb{\ell}{\mhole{a}}\) 
    \;|\; \(\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}\)
    \;|\; \(\letstmt{x}{\mhole{a}}{\mhole{t}}\)
    \;|\; \(\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}\)
    \;|\; \(\ewhere{\mhole{t}}{\mhole{L}}\)

    <\(\mhole{L}\)> ::= \(\lhole{W}\) \;|\; \(\cdot\) \;|\; \(\lwbranch{\ell}{x: A}{\mhole{t}}, \mhole{L}\)

    <\(H\)> ::= \(\cdot\) 
    \;|\; \(H, \tyhole{\lhole{X}}{\Gamma}{p}{A}{q}\)
    \;|\; \(H, \blkhole{\lhole{T}}{\Gamma}{\ms{L}}\)
    \;|\; \(H, \cfghole{\lhole{W}}{\ms{L}}{\ms{K}}\)
  \end{grammar}
  \caption{Grammar for \isotopessa terms and blocks with holes}
  \Description{Grammar for isotope-SSA blocks with holes}
  \label{fig:blocks-with-holes-grammar}
\end{figure}

\begin{figure}
  \begin{gather*}    
    \prftree[r]{\rle{hole-tm}}
      {\tyhole{\lhole{X}}{\Delta}{p}{A}{q}}
      {\cwk{\Gamma}{\Delta}}
      {\mhasty{H}{\Gamma}{p}{\lhole{X}}{A}{q}} \qquad
    \prftree[r]{\rle{var}}
      {\cwk{\Gamma}{\thyp{x}{A}{q}}}
      {\mhasty{H}{\Gamma}{p}{x}{A}{q}} \qquad
    \prftree[r]{\rle{app}}
      {f \in \mc{I}_p^q(A, B)}
      {\mhasty{H}{\Gamma}{1}{\mhole{a}}{A}{q}}
      {\mhasty{H}{\Gamma}{p}{f\;\mhole{a}}{B}{q}} \\
    \prftree[r]{\rle{pair}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\mhasty{H}{\Delta}{1}{\mhole{a}}{A}{q}}
      {\mhasty{H}{\Xi}{1}{\mhole{b}}{B}{q}}
      {\mhasty{H}{\Gamma}{p}{(\mhole{a}, \mhole{b})}{A \otimes B}{q}} \\
    \prftree[r]{\rle{unit}}
      {\cwk{\Gamma}{\cdot}}
      {\mhasty{H}{\Gamma}{p}{()}{\mb{1}}{q}} \qquad
    \prftree[r]{\rle{true}}
      {\cwk{\Gamma}{\cdot}}
      {\mhasty{H}{\Gamma}{p}{\ctt}{\mb{2}}{q}} \qquad
    \prftree[r]{\rle{false}}
      {\cwk{\Gamma}{\cdot}}
      {\mhasty{H}{\Gamma}{p}{\cff}{\mb{2}}{q}} \\
    \prftree[r]{\rle{let}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\mhasty{H}{\Delta, \thyp{x}{A}{}}{p}{\mhole{e}}{B}{q}}
      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{q}}
      {\mhasty{H}{\Gamma}{p}{\letexpr{x}{\mhole{a}}{\mhole{e}}}{B}{q}} \qquad
    \prftree[r]{\rle{blk}}
      {\mhaslb{H}{\Gamma}{\mhole{t}}{\llhyp{\ell}{\cdot}{A}}}
      {\mhasty{H}{\Gamma}{0}{\lbsplice{\ell}{A}{\mhole{t}}}{A}{\varnothing}} \\
    \prftree[r]{\rle{let2}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\mhasty{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{p}{\mhole{e}}{C}{q}}
      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A \otimes B}{q}}
      {\mhasty{H}{\Gamma}{p}{\letexpr{(x, y)}{\mhole{a}}{\mhole{e}}}{C}{q}}
  \end{gather*}
  \caption{Typing rules for \isotopessa terms with holes}
  \Description{Typing rules for isotope-SSA terms with holes}
  \label{fig:terms-with-holes-typing}
\end{figure}

\begin{figure}
  \begin{gather*}    
    \prftree[r]{\rle{hole-blk}}
      {\blkhole{\lhole{T}}{\Delta}{\ms{L}} \in H}
      {\cwk{\Gamma}{\Delta}}
      {\mhaslb{H}{\Gamma}{\lhole{T}}{\ms{L}}} 
      \qquad
    \prftree[r]{\rle{br}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}}
      {\mhasty{H}{\Xi}{0}{\mhole{a}}{A}{\tint}}
      {\mhaslb{H}{\Gamma}{\lbrb{\ell}{\mhole{a}}}{\ms{L}}} 
      \\
    \prftree[r]{\rle{ite}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\mhasty{H}{\Delta}{0}{\mhole{e}}{\mb{2}}{\tint}}
      {\mhaslb{H}{\Xi}{\mhole{s}}{\ms{L}}}
      {\mhaslb{H}{\Xi}{\mhole{t}}{\ms{L}}}
      {\mhaslb{H}{\Gamma}{\ite{\mhole{e}}{\mhole{s}}{\mhole{t}}}{\ms{L}}} \\
    \prftree[r]{\rle{let-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\mhaslb{H}{\Delta, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A}{q}}
      {\mhaslb{H}{\Gamma}{\letstmt{x}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
    \prftree[r]{\rle{let2-blk}}
      {\csplits{\Gamma}{\Delta}{\Xi}}
      {\mhaslb{H}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}{\mhole{t}}{\ms{L}}}
      {\mhasty{H}{\Xi}{1}{\mhole{a}}{A \otimes B}{q}}
      {\mhaslb{H}{\Gamma}{\letstmt{(x, y)}{\mhole{a}}{\mhole{t}}}{\ms{L}}} \\
    \prftree[r]{\rle{where}}
      {\mhaslb{H}{\Gamma}{\mhole{t}}{\ms{L}}}
      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}}}
      {\mhaslb{H}{\Gamma}{\ewhere{\mhole{t}}{\mhole{L}}}{\ms{K}}} 
      \qquad
    \prftree[r]{\rle{hole-cfg}}
      {\cfghole{\lhole{W}}{\ms{L}}{\ms{K}} \in H}
      {\lwk{\ms{M}}{\ms{K}}}
      {\mlhaslb{\ms{L}}{H}{\lhole{W}}{\ms{K}}}
      \\
    \prftree[r]{\rle{nil-br}}
      {\mlhaslb{H}{\ms{L}}{\cdot}{\ms{L}}} \qquad
    \prftree[r]{\rle{cons-br}}
      {\mlhaslb{H}{\ms{L}}{\mhole{L}}{\ms{K}, \llhyp{\ell}{\Gamma}{A}}}
      {\mhaslb{H}{\Gamma, \thyp{x}{A}{}}{\mhole{t}}{\ms{L}}}
      {\mlhaslb{H}{\ms{L}}{\mhole{L}, \lwbranch{\ell}{x: A}{\mhole{t}}}{\ms{K}}}
  \end{gather*}
  \caption{Typing rules for \isotopessa blocks with holes}
  \Description{Typing rules for isotope-SSA blocks with holes}
  \label{fig:blocks-with-holes-typing}
\end{figure}

\begin{figure}
  \begin{gather*}
    \boxed{H: \ms{Hole} \to \ms{MExpr} \sqcup \ms{MBlock}}
    \\
    \sorry
  \end{gather*}
  \caption{Typing rules for \isotopessa rewrites}
  \Description{Typing rules for isotope-SSA rewrites}
  \label{fig:rewrite-typing}
\end{figure}

\begin{theorem}[Rewriting]
  Given \(\isrw{\mc{H}}{H}{I}\), we have
  \begin{itemize}
    \item If \(\mhasty{H}{\Gamma}{p}{a}{A}{q}\), then \(\mhasty{I}{\Gamma}{p}{[\mc{H}]a}{A}{q}\)
    \item If \(\mhaslb{H}{\Gamma}{t}{\ms{L}}\), then \(\mhaslb{I}{\Gamma}{[\mc{H}]t}{\ms{L}}\)
  \end{itemize}
  Furthermore, we have that 
  \begin{itemize}
    \item \(\mhasty{\cdot}{\Gamma}{p}{a}{A}{q} \iff \hasty{\Gamma}{p}{a}{A}{q}\)
    \item \(\mhaslb{\cdot}{\Gamma}{t}{\ms{L}} \iff \haslb{\Gamma}{t}{\ms{L}}\)
  \end{itemize}
\end{theorem}

@@

\TODO{rewriting substitution + weakening?}
